Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-15 MS_thesis.tex
More comprehensive commenting
Add some fancy formatting: Remove page no.s from Part titles, make sure appendix starts on a fresh odd page
add a \clearpage between main text and appendices
Headers on odd pages of appendix now include the appendix index (A,B,C,..)
Separate discussion into separate technical and non-technical pchapters
Slightly rephrase abstract

\documentclass[twoside, 12pt]{iiser-thesis-modified}

@@

%my own list of packages
\input{head}

%%%%%%%%

%Add fancy headers with chapter/section titles
\usepackage{fancyhdr}
\setlength{\headsep}{1.8in} %add spacing between header and body text

%%%%%%%%

@@

\usepackage{setspace}
\onehalfspacing

\setcounter{tocdepth}{3}

@@

%%%%%%%%%%%%
% Document %
%%%%%%%%%%%%

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis through fields such as theoretical population genetics, evolutionary game theory, and quantitative genetics. Historically, these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolving populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. Along with recovering the effects of the standard evolutionary forces of selection, mutation, and drift, these SDEs also reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite populations and has been largely overlooked in standard formulations of evolution. Noise-induced selection can reverse the direction of evolution predicted by infinite-population frameworks, with implications for simulation studies and real world populations. Well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. Finally, I extend these ideas to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite fluctuating populations and predicts a new evolutionary force unique to such populations.
}

\acknowledgments{Not more than 250 words}
\allowdisplaybreaks %allow align envts to go across pages

@@

\pagestyle{plain}
\thesisfront
\listoftables
\addcontentsline{toc}{chapter}{List of Tables}
\listoffigures
\addcontentsline{toc}{chapter}{List of Figures}

\part{Background}

@@

\chapter{Introduction}
\input{ind_files/01_Introduction}

\part{Theory}\label{part_theory}
\chapter{Mathematical background and an expository example}\label{chap_math_background}
\input{ind_files/02_math_background}
\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes}
\chapter{Stochastic field equations for the evolution of quantitative traits}\label{chap_infD_processes}
\input{ind_files/04_field_eqns}

\part{Major Takeaways \& Discussion}\label{part_summary}
\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/05_unification_discussion}

%You need the below commands to elevate the appendix to 'Part' level in the TOC
%from https://tex.stackexchange.com/a/286452
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

%make sure the title is 'Appendix' and not 'Chapter'
\renewcommand{\chaptername}{Appendix} 
\chapter{Deriving the Fokker-Planck equations for It\^{o} SDEs}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}

\cleardoublepage
\phantomsection
\printbibliography
\addcontentsline{toc}{part}{References}
\end{document}

>>>>

\documentclass[twoside, 12pt,lot,lof]{iiser-thesis-modified} %Use the custom .cls file we have

@@

%%%%%%%%%%%%%%%%%%%

%My own list of packages and commands, for convenience
\input{head}

%%%%%%%%%%%%%%%%%%%

\usepackage{fancyhdr} %For fancy page headers
\setlength{\headsep}{1.8in} %add spacing between header and body text

%%%%%%%%%%%%%%%%%%%

@@

%1.5 line spacing (as requested by IISER)
\usepackage{setspace}
\onehalfspacing

%Decide upto what depth the table of contents should list: subsection, subsubsection, etc
\setcounter{tocdepth}{3}

@@

%%%%%%%%%%%%%%%%%%%
%  Abstract and   %
%  other details  %
%%%%%%%%%%%%%%%%%%%

@@

%The abstract goes here. IISER guidelines say it should be <= 250 words.
\thesisabstract{
Population biology is built on a strong mathematical foundation developed during the Modern Synthesis and encapsulated in fields such as population genetics, evolutionary game theory, and quantitative genetics. Historically, many of these formalisms have often worked with infinite populations, ignoring the effects of demographic stochasticity. Finite population models in population and quantitative genetics usually assume a fixed population size and are of limited applicability in the real world, where population sizes routinely fluctuate. In this thesis, I use ideas from statistical physics to analytically describe evolution in finite, fluctuating populations from biological first principles. Starting from a density-dependent `birth-death process’ describing an arbitrary closed population of individuals with discrete traits, I derive a set of stochastic differential equations (SDEs) for how trait frequencies change over time. Along with recovering the effects of the standard evolutionary forces of selection, mutation, and drift, these SDEs also reveal a new directional evolutionary force, `noise-induced selection’, that is particular to finite, fluctuating populations and has therefore been largely overlooked in standard formulations of evolution. Noise-induced selection can cause deviations from neutrality and can reverse the direction of evolution predicted by infinite-population frameworks, with implications for simulation studies and real-world populations. Well-known results such as the replicator-mutator equation and Fisher’s fundamental theorem are recovered in the infinite population limit. Finally, I extend these ideas to one-dimensional quantitative traits through a `stochastic field theory’ that yields frameworks such as Kimura's continuum-of-alleles and Lande's gradient dynamics in the infinite population limit. My work thus generalizes the formal structures of population biology to finite, fluctuating populations and predicts a new evolutionary force unique to such populations.
}

%The acknowledgements go here. IISER guidelines say it should be <= 250 words.
\acknowledgments{
Not more than 250 words
}
\allowdisplaybreaks %allow align envts to go across pages

%%%%%%%%%%%%%%%%%%%
%  The Document   %
%%%%%%%%%%%%%%%%%%%

@@

\pagestyle{plain} %Don't make the initial frontmatter fancy
\thesisfront
%\listoftables
%\addcontentsline{toc}{chapter}{List of Tables}
%\listoffigures
%\addcontentsline{toc}{chapter}{List of Figures}

%wrap the declaration of every new \part in the \clear and the group to ensure that the fresh page created by \part does not have a page number.
% This hack-ey workaround is from https://tex.stackexchange.com/a/191189 and is needed because \part defines \pagestyle WITHIN the definition itself and so you cannot use \thispagestyle to mess with it
\cleardoublepage
\begingroup
\makeatletter
\let\ps@plain\ps@empty
\part{Background}
\endgroup

@@

%From here on, we just need to import TeX files from the ind_files folder to include the actual content of the thesis

%Intro
\chapter{Introduction}
\input{ind_files/01_Introduction}

%The 'meat': Methods, results, theory, etc.
\cleardoublepage
\begingroup
\makeatletter
\let\ps@plain\ps@empty
\part{Theory}\label{part_theory}
\endgroup

\chapter{Mathematical background and an expository example}\label{chap_math_background}
\input{ind_files/02_math_background}

\chapter{Population dynamics from stochastic first principles}\label{chap_BD}
\input{ind_files/03_BD_processes}

\chapter{Stochastic field equations for the evolution of quantitative traits}\label{chap_infD_processes}
\input{ind_files/04_field_eqns}

%Discussion and summary
\cleardoublepage
\begingroup
\makeatletter
\let\ps@plain\ps@empty
\part{Major Takeaways \& Discussion}\label{part_summary}
\endgroup

\chapter{A unified view of population dynamics}\label{chap_unification}
\input{ind_files/05_unification_discussion}

\chapter{Discussion and Outlook}\label{sec_disc}
\input{ind_files/06_nontechnical_discussion}

%Add a new page after all the chapters are done, just to separate the main text from the appendices. If adding a new page results in you being on an even page, add another new page, just to ensure that you always begin with text on an odd page (useful if you intend to print the thesis at some point).
\clearpage
\null\thispagestyle{plain}\newpage
\ifodd\value{page}\else\null\thispagestyle{plain}\newpage\fi

%You need the below commands to add a 'Part' level reference to the Appendices in the TOC
%from https://tex.stackexchange.com/a/286452
\cleardoublepage
\phantomsection
\appendix
\addcontentsline{toc}{part}{Appendices}

%make sure the title of the appendices say 'Appendix' and not 'Chapter' even though we declare them at `chapter level' internally
\renewcommand{\chaptername}{Appendix}

%For the headers of appendices (on odd pages), preface the appendix title with 'Appendix X', where X is the Appendix index (A,B,C,...)
\fancyhead[CO]{  \rule[-4ex]{0pt}{4ex}\footnotesize\itshape{\chaptername~\thechapter: \nouppercase{\leftmark}}}

%Import the Appendices
\chapter{From It\^{o} to Fokker-Planck}\label{App_SDE_FPE}
\input{ind_files/99_Appendix_A}
\chapter{Deriving stochastic trait frequency dynamics using It\^{o}'s formula}\label{App_density_to_freq}
\input{ind_files/99_Appendix_B}
\chapter{A Price-like equation for the variance of a type-level quantity}\label{App_stoch_var_eqns}
\input{ind_files/99_Appendix_C}
\chapter{Some Examples}\label{App_examples}
\input{ind_files/99_Appendix_D_examples}

%Add the bibliography and make an entry in the table of contents
\cleardoublepage
\phantomsection
\printbibliography
\addcontentsline{toc}{part}{References}

%And we're done :)
\end{document}
